---
source: Rmd
title: Next steps
teaching: 45
exercises: 45
---

```{r, include=FALSE}
```

::::::::::::::::::::::::::::::::::::::: objectives

- Introduce the Bioconductor project.
- Introduce the notion of data containers.
- Give an overview of the `SummarizedExperiment`, extensively used in
  omics analyses.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What is a `SummarizedExperiment`?
- What is Bioconductor?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Next steps

```{r, echo=FALSE, message=FALSE}
library("tidyverse")
```

Data in bioinformatics is often complex.  To deal with this,
developers define specialised data containers (termed classes) that
match the properties of the data they need to handle.

This aspect is central to the **Bioconductor**[^Bioconductor] project
which uses the same **core data infrastructure** across packages. This
certainly contributed to Bioconductor's success. Bioconductor package
developers are advised to make use of existing infrastructure to
provide coherence, interoperability, and stability to the project as a
whole.

[^Bioconductor]: The [Bioconductor](https://www.bioconductor.org) was
    initiated by Robert Gentleman, one of the two creators of the R
    language. Bioconductor provides tools dedicated to omics data
    analysis. Bioconductor uses the R statistical programming language
    and is open source and open development.

To illustrate such an omics data container, we'll present the
`SummarizedExperiment` class.

## SummarizedExperiment

The figure below represents the anatomy of the SummarizedExperiment class.

```{r SE, echo=FALSE, out.width="80%"}
knitr::include_graphics("https://uclouvain-cbio.github.io/WSBIM1322/figs/SE.svg")
```

Objects of the class SummarizedExperiment contain :

- **One (or more) assay(s)** containing the quantitative omics data
  (expression data), stored as a matrix-like object. Features (genes,
  transcripts, proteins, ...) are defined along the rows, and samples
  along the columns.

- A **sample metadata** slot containing sample co-variates, stored as a
  data frame. Rows from this table represent samples (rows match exactly the
  columns of the expression data).

- A **feature metadata** slot containing feature co-variates, stored as
  a data frame. The rows of this data frame match exactly the rows of the
  expression data.

The coordinated nature of the SummarizedExperiment guarantees that
during data manipulation, the dimensions of the different slots will
always match (i.e the columns in the expression data and then rows in
the sample metadata, as well as the rows in the expression data and
feature metadata) during data manipulation. For example, if we had to
exclude one sample from the assay, it would be automatically removed
from the sample metadata in the same operation.

The metadata slots can grow additional co-variates
(columns) without affecting the other structures.

### Creating a SummarizedExperiment

In order to create a `SummarizedExperiment`, we will create the
individual components, i.e the count matrix, the sample and gene
metadata from csv files. These are typically how RNA-Seq data are
provided (after raw data have been processed).

```{r, echo=FALSE, message=FALSE}
rna <- read_csv("data/rnaseq.csv")

## count matrix
counts <- rna %>%
  select(gene, sample, expression) %>%
  pivot_wider(names_from = sample,
              values_from = expression)

## convert to matrix and set row names
count_matrix <- counts %>% select(-gene) %>% as.matrix()
rownames(count_matrix) <- counts$gene

## sample annotation
sample_metadata <- rna %>%
  select(sample, organism, age, sex, infection, strain, time, tissue, mouse)

## remove redundancy
sample_metadata <- unique(sample_metadata)

## gene annotation
gene_metadata <- rna %>%
  select(gene, ENTREZID, product, ensembl_gene_id, external_synonym,
         chromosome_name, gene_biotype, phenotype_description,
         hsapiens_homolog_associated_gene_name)

# remove redundancy
gene_metadata <- unique(gene_metadata)

## write to csv
write.csv(count_matrix, file = "data/count_matrix.csv")
write.csv(gene_metadata, file = "data/gene_metadata.csv", row.names = FALSE)
write.csv(sample_metadata, file = "data/sample_metadata.csv", row.names = FALSE)
```

- **An expression matrix**: we load the count matrix, specifying that
  the first columns contains row/gene names, and convert the
  `data.frame` to a `matrix`. You can download it
  [here](https://carpentries-incubator.github.io/bioc-intro/data/count_matrix.csv).

```{r}
count_matrix <- read.csv("data/count_matrix.csv",
                         row.names = 1) %>%
    as.matrix()

count_matrix[1:5, ]
dim(count_matrix)
```

- **A table describing the samples**, available
  [here](https://carpentries-incubator.github.io/bioc-intro/data/sample_metadata.csv).

```{r}
sample_metadata <- read.csv("data/sample_metadata.csv")
sample_metadata
dim(sample_metadata)
```

- **A table describing the genes**, available
  [here](https://carpentries-incubator.github.io/bioc-intro/data/gene_metadata.csv).

```{r}
gene_metadata <- read.csv("data/gene_metadata.csv")
gene_metadata[1:10, 1:4]
dim(gene_metadata)
```

We will create a `SummarizedExperiment` from these tables:

- The count matrix that will be used as the **`assay`**

- The table describing the samples will be used as the **sample
  metadata** slot

- The table describing the genes will be used as the **features
  metadata** slot

To do this we can put the different parts together using the
`SummarizedExperiment` constructor:

```{r, message=FALSE}
## BiocManager::install("SummarizedExperiment")
library("SummarizedExperiment")
```

First, we make sure that the samples are in the same order in the
count matrix and the sample annotation, and the same for the genes in
the count matrix and the gene annotation.

```{r}
stopifnot(rownames(count_matrix) == gene_metadata$gene)
stopifnot(colnames(count_matrix) == sample_metadata$sample)
```

```{r}
se <- SummarizedExperiment(assays = list(counts = count_matrix),
                           colData = sample_metadata,
                           rowData = gene_metadata)
se
```

### Saving data

Exporting data to a spreadsheet, as we did in a previous episode, has
several limitations, such as those described in the first chapter
(possible inconsistencies with `,` and `.` for decimal separators and
lack of variable type definitions). Furthermore, exporting data to a
spreadsheet is only relevant for rectangular data such as dataframes
and matrices.

A more general way to save data, that is specific to R and is
guaranteed to work on any operating system, is to use the `saveRDS`
function. Saving objects like this will generate a binary
representation on disk (using the `rds` file extension here), which
can be loaded back into R using the `readRDS` function.

```{r, eval=FALSE}
saveRDS(se, file = "data_output/se.rds")
rm(se)
se <- readRDS("data_output/se.rds")
head(se)
```

To conclude, when it comes to saving data from R that will be loaded
again in R, saving and loading with `saveRDS` and `readRDS` is the
preferred approach. If tabular data need to be shared with somebody
that is not using R, then exporting to a text-based spreadsheet is a
good alternative.

Using this data structure, we can access the expression matrix with
the `assay` function:

```{r}
head(assay(se))
dim(assay(se))
```

We can access the sample metadata using the `colData` function:

```{r}
colData(se)
dim(colData(se))
```

We can also access the feature metadata using the `rowData` function:

```{r}
head(rowData(se))
dim(rowData(se))
```

### Subsetting a SummarizedExperiment

SummarizedExperiment can be subset just like with data frames, with
numerics or with characters of logicals.

Below, we create a new instance of class SummarizedExperiment that
contains only the 5 first features for the 3 first samples.

```{r}
se1 <- se[1:5, 1:3]
se1
```

```{r}
colData(se1)
rowData(se1)
```

We can also use the `colData()` function to subset on something from
the sample metadata or the `rowData()` to subset on something from the
feature metadata.  For example, here we keep only miRNAs and the non
infected samples:

```{r}
se1 <- se[rowData(se)$gene_biotype == "miRNA",
          colData(se)$infection == "NonInfected"]
se1
assay(se1)
colData(se1)
rowData(se1)
```

<!--For the following exercise, you should download the SE.rda object
(that contains the `se` object), and open the file using the 'load()'
function.-->

<!-- ```{r, eval = FALSE, echo = FALSE} -->

<!-- download.file(url = "https://raw.githubusercontent.com/UCLouvain-CBIO/bioinfo-training-01-intro-r/master/data/SE.rda", -->

<!--               destfile = "data/SE.rda") -->

<!-- load("data/SE.rda") -->

<!-- ``` -->

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge

Extract the gene expression levels of the 3 first genes in samples
at time 0 and at time 8.

:::::::::::::::  solution

## Solution

```{r, purl=FALSE}
assay(se)[1:3, colData(se)$time != 4]

# Equivalent to
assay(se)[1:3, colData(se)$time == 0 | colData(se)$time == 8]
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge

Verify that you get the same values using the long `rna` table.

:::::::::::::::  solution

## Solution

```{r, purl=FALSE}
rna |>
    filter(gene %in% c("Asl", "Apod", "Cyd2d22")) |>
    filter(time != 4) |> select(expression)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

The long table and the `SummarizedExperiment` contain the same
information, but simply structured differently. Both have they
advantages: the former is a good fit for the `tidyverse` packages,
while the latter is the preferred structure for many bioinformatics and
statistical processing steps, such as a typical RNA-Seq analyses using
the `DESeq2` package.

#### Adding variables to metadata

We can also add information to the metadata.
Suppose that you want to add the center where the samples were collected...

```{r}
colData(se)$center <- rep("University of Illinois", nrow(colData(se)))
colData(se)
```

This illustrates that the metadata slots can grow indefinitely without
affecting the other structures!

### tidySummarizedExperiment

You may be wondering, can we use tidyverse commands to interact with
`SummarizedExperiment` objects. The answer is yes, we can with the
`tidySummarizedExperiment` package.

Remember what our SummarizedExperiment object looks like.

```{r, message=FALSE}
se
```

Load `tidySummarizedExperiment` and then take a look at the se object
again.

```{r, message=FALSE}
#BiocManager::install("tidySummarizedExperiment")
library("tidySummarizedExperiment")

se
```

It's still a `SummarizedExperiment` object, so maintains the efficient
structure, but now we can view it as a tibble. Note the first line of
the output says this, it's a `SummarizedExperiment`\-`tibble`
abstraction. We can also see in the second line of the output the
number of transcripts and samples.

If we want to revert to the standard `SummarizedExperiment` view we
can do that.

```{r}
options("restore_SummarizedExperiment_show" = TRUE)
se
```

But here we will use the tibble view.

```{r}
options("restore_SummarizedExperiment_show" = FALSE)
se
```

We can now use tidyverse commands to interact with the
SummarizedExperiment object.

We can use `filter` to filter for rows using a condition e.g. to view
all rows for one sample.

```{r}
se %>% filter(.sample == "GSM2545336")
```

We can use `select` to specify columns we want to view.

```{r}
se %>% select(.sample)
```

We can use `mutate` to add metadata info.

```{r}
se %>% mutate(center = "Heidelberg University")
```

We can also combine commands with the tidyverse pipe `%>%`. For
example, we could combine `group_by` and `summarise` to get the total
counts for each sample.

```{r}
se %>%
    group_by(.sample) %>%
    summarise(total_counts=sum(counts))
```

We can treat the tidy SummarizedExperiment object as a normal tibble
for plotting.

Here we plot the distribution of counts per sample.

```{r tidySE-plot}
se %>%
    ggplot(aes(counts + 1, group=.sample, color=infection)) +
    geom_density() +
    scale_x_log10() +
    theme_bw()
```

For more information on tidySummarizedExperiment, see the package
website
[here](https://stemangiola.github.io/tidySummarizedExperiment/).

**Take-home message**

- `SummarizedExperiment` represents an efficient way to store and
  handle omics data.

- They are used in many Bioconductor packages.

If you follow the next training focused on RNA sequencing analysis,
you will learn to use the Bioconductor `DESeq2` package to do some
differential expression analyses.  The whole analysis of the `DESeq2`
package is handled in a `SummarizedExperiment`.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Bioconductor is a project provide support and packages for the
  comprehension of high high-throughput biology data.
- A `SummarizedExperiment` is a type of object useful to store and
  manage high-throughput omics data.

::::::::::::::::::::::::::::::::::::::::::::::::::
