---
source: Rmd
title: Next steps
teaching: 45
exercises: 45
---

```{r, include=FALSE}
```

::::::::::::::::::::::::::::::::::::::: objectives

- Introduce the Bioconductor project.
- Introduce the notion of data containers.
- Give an overview of the `SummarizedExperiment`, extensively used in
  omics analyses.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- What is a `SummarizedExperiment`?
- What is Bioconductor?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Next steps

```{r, echo=FALSE, message=FALSE}
library("tidyverse")
```

Data in bioinformatics is often complex.  Para lidar com este problema, os programadores de
definem contentores de dados especializados (denominados de classes) que correspondem às propriedades dos dados que necessitam ser manipulados.

Este aspeto é fundamental para o projeto do **Bioconductor**[^Bioconductor] que utiliza a mesma **infraestrutura de dados central** em todos os pacotes. Esta estrutura contribuiu certamente para o sucesso do Bioconductor. Os programadores de pacotes do Bioconductor
são aconselhados a utilizar a infraestrutura existente para
proporcionar coerência, interoperabilidade e estabilidade ao projeto como um todo.

[^Bioconductor]: The [Bioconductor](https://www.bioconductor.org) was
    initiated by Robert Gentleman, one of the two creators of the R
    language. O Bioconductor fornece ferramentas dedicadas a análise de dados ômicos. O Bioconductor utiliza a linguagem de programação estatística R e tem o código e o desenvolvimento aberto.

Para ilustrar um contêiner de dados ômicos, apresentaremos a classe
`SummarizedExperiment`.

## SummarizedExperiment

The figure below represents the anatomy of the SummarizedExperiment class.

```{r SE, echo=FALSE, out.width="80%"}
knitr::include_graphics("https://uclouvain-cbio.github.io/WSBIM1322/figs/SE.svg")
```

Objects of the class SummarizedExperiment contain :

- **One (or more) assay(s)** containing the quantitative omics data
  (expression data), stored as a matrix-like object. Features (genes,
  transcripts, proteins, ...) are defined along the rows, and samples
  along the columns.

- A **sample metadata** slot containing sample co-variates, stored as a
  data frame. As linhas desta tabela representam amostras (as linhas correspondem exatamente às colunas
  dos dados de expressão gênica).

- A **feature metadata** slot containing feature co-variates, stored as
  a data frame. As linhas desta estrutura de dados correspondem exatamente às linhas dos dados da expressão.

A natureza coordenada do `SummarizedExperiment` garante que
durante a manipulação dos dados, as dimensões dos diferentes compartimentos serão
sempre correspondentes (por exemplo as colunas nos dados de expressão e, em seguida, as linhas nos metadados da amostra, bem como as linhas nos dados de expressão e
metadados das variáveis) durante a manipulação dos dados. Por exemplo, se tivéssemos que
excluir uma amostra do ensaio, esta seria automaticamente removida
dos metadados da amostra na mesma operação.

Os compartimentos de metadados podem aumentar as co-variáveis adicionais
(colunas) sem afetar as outras estruturas.

### Creating a SummarizedExperiment

In order to create a `SummarizedExperiment`, we will create the
individual components, i.e the count matrix, the sample and gene
metadata from csv files. Normalmente, é assim que os dados de RNA-Seq são
fornecidos (depois dos dados brutos terem sido processados).

```{r, echo=FALSE, message=FALSE}
rna <- read_csv("data/rnaseq.csv")

## count matrix
counts <- rna |>
  select(gene, sample, expression) |>
  pivot_wider(names_from = sample,
              values_from = expression)

## convert to matrix and set row names
count_matrix <- counts |>
  select(-gene) |>
  as.matrix()
rownames(count_matrix) <- counts$gene

## sample annotation
sample_metadata <- rna |>
  select(sample, organism, age, sex, infection, strain, time, tissue, mouse)

## remove redundancy
sample_metadata <- unique(sample_metadata)

## gene annotation
gene_metadata <- rna |>
  select(gene, ENTREZID, product, ensembl_gene_id, external_synonym,
         chromosome_name, gene_biotype, phenotype_description,
         hsapiens_homolog_associated_gene_name)

# remove redundancy
gene_metadata <- unique(gene_metadata)

## write to csv
write.csv(count_matrix, file = "data/count_matrix.csv")
write.csv(gene_metadata, file = "data/gene_metadata.csv", row.names = FALSE)
write.csv(sample_metadata, file = "data/sample_metadata.csv", row.names = FALSE)
```

- **An expression matrix**: we load the count matrix, specifying that
  the first columns contains row/gene names, and convert the
  `data.frame` to a `matrix`. You can download it
  [here](https://carpentries-incubator.github.io/bioc-intro/data/count_matrix.csv).

```{r}
count_matrix <- read.csv("data/count_matrix.csv",
                         row.names = 1) |>
    as.matrix()

count_matrix[1:5, ]
dim(count_matrix)
```

- **A table describing the samples**, available
  [here](https://carpentries-incubator.github.io/bioc-intro/data/sample_metadata.csv).

```{r}
sample_metadata <- read.csv("data/sample_metadata.csv")
sample_metadata
dim(sample_metadata)
```

- **A table describing the genes**, available
  [here](https://carpentries-incubator.github.io/bioc-intro/data/gene_metadata.csv).

```{r}
gene_metadata <- read.csv("data/gene_metadata.csv")
gene_metadata[1:10, 1:4]
dim(gene_metadata)
```

We will create a `SummarizedExperiment` from these tables:

- The count matrix that will be used as the **`assay`**

- The table describing the samples will be used as the **sample
  metadata** slot

- The table describing the genes will be used as the **features
  metadata** slot

To do this we can put the different parts together using the
`SummarizedExperiment` constructor:

```{r, message=FALSE, warning=FALSE}
## BiocManager::install("SummarizedExperiment")
library("SummarizedExperiment")
```

First, we make sure that the samples are in the same order in the
count matrix and the sample annotation, and the same for the genes in
the count matrix and the gene annotation.

```{r}
stopifnot(rownames(count_matrix) == gene_metadata$gene)
stopifnot(colnames(count_matrix) == sample_metadata$sample)
```

```{r}
se <- SummarizedExperiment(assays = list(counts = count_matrix),
                           colData = sample_metadata,
                           rowData = gene_metadata)
se
```

### Saving data

Exporting data to a spreadsheet, as we did in a previous episode, has
several limitations, such as those described in the first chapter
(possible inconsistencies with `,` and `.` for decimal separators and
lack of variable type definitions). Além disso, a exportação dos dados para uma spreadsheet
só é relevante para dados retangulares, tais como data.frames
e matrizes.

Uma forma mais geral de guardar dados, que é específica do R e é
garantida para funcionar em qualquer sistema operativo, é utilizar a função `saveRDS`. Guardar objetos como este irá gerar uma representação binária
no disco (usando a extensão de arquivo `rds` aqui), que
pode ser carregada de volta para o R usando a função `readRDS`.

```{r, eval=FALSE}
saveRDS(se, file = "data_output/se.rds")
rm(se)
se <- readRDS("data_output/se.rds")
head(se)
```

To conclude, when it comes to saving data from R that will be loaded
again in R, saving and loading with `saveRDS` and `readRDS` is the
preferred approach. Se os dados tabulares tiverem de ser partilhados com alguém
que não utilize o R, então a exportação para spreadsheet baseada em texto é uma
boa alternativa.

Using this data structure, we can access the expression matrix with
the `assay` function:

```{r}
head(assay(se))
dim(assay(se))
```

We can access the sample metadata using the `colData` function:

```{r}
colData(se)
dim(colData(se))
```

We can also access the feature metadata using the `rowData` function:

```{r}
head(rowData(se))
dim(rowData(se))
```

### Subsetting a SummarizedExperiment

SummarizedExperiment can be subset just like with data frames, with
numerics or with characters of logicals.

Abaixo, criamos uma nova instância da classe SummarizedExperiment que contém apenas as 5 primeiras variáveis para as 3 primeiras amostras.

```{r}
se1 <- se[1:5, 1:3]
se1
```

```{r}
colData(se1)
rowData(se1)
```

We can also use the `colData()` function to subset on something from
the sample metadata or the `rowData()` to subset on something from the
feature metadata.  For example, here we keep only miRNAs and the non
infected samples:

```{r}
se1 <- se[rowData(se)$gene_biotype == "miRNA",
          colData(se)$infection == "NonInfected"]
se1
assay(se1)
colData(se1)
rowData(se1)
```

<!--For the following exercise, you should download the SE.rda object
(that contains the `se` object), and open the file using the 'load()'
function.-->

<!-- ```{r, eval = FALSE, echo = FALSE} -->

<!-- download.file(url = "https://raw.githubusercontent.com/UCLouvain-CBIO/bioinfo-training-01-intro-r/master/data/SE.rda", -->

<!--               destfile = "data/SE.rda") -->

<!-- load("data/SE.rda") -->

<!-- ``` -->

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge

Extract the gene expression levels of the 3 first genes in samples
at time 0 and at time 8.

:::::::::::::::  solution

## Solution

```{r, purl=FALSE}
assay(se)[1:3, colData(se)$time != 4]

# Equivalent to
assay(se)[1:3, colData(se)$time == 0 | colData(se)$time == 8]
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge

Verify that you get the same values using the long `rna` table.

:::::::::::::::  solution

## Solution

```{r, purl=FALSE}
rna |>
    filter(gene %in% c("Asl", "Apod", "Cyd2d22")) |>
    filter(time != 4) |>
    select(expression)
```

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::

The long table and the `SummarizedExperiment` contain the same
information, but are simply structured differently. Cada abordagem tem as suas
próprias vantagens: a primeira adequa-se bem aos pacotes `tidyverse`,
enquanto a segunda é a estrutura preferida para muitas etapas de processamento bioinformático e
estatístico. Por exemplo, uma análise típica de RNA-Seq utilizando
o pacote `DESeq2`.

#### Adding variables to metadata

We can also add information to the metadata.
Suponhamos que se pretende acrescentar o local onde as amostras foram recolhidas...

```{r}
colData(se)$center <- rep("University of Illinois", nrow(colData(se)))
colData(se)
```

This illustrates that the metadata slots can grow indefinitely without
affecting the other structures!

### tidySummarizedExperiment

You may be wondering, can we use tidyverse commands to interact with
`SummarizedExperiment` objects? A resposta é sim, podemos fazê-lo com o pacote
`tidySummarizedExperiment`.

Remember what our SummarizedExperiment object looks like:

```{r, message=FALSE}
se
```

Load `tidySummarizedExperiment` and then take a look at the se object
again.

```{r, message=FALSE}
#BiocManager::install("tidySummarizedExperiment")
library("tidySummarizedExperiment")

se
```

It's still a `SummarizedExperiment` object, so maintains the efficient
structure, but now we can view it as a tibble. Repare que na primeira linha do output diz isto:
`SummarizedExperiment`\-`tibble`
abstraction. Também podemos ver na segunda linha do output o
número de transcrições e amostras.

Se quisermos, podemos reverter para a visualização padrão do `SummarizedExperiment`.

```{r}
options("restore_SummarizedExperiment_show" = TRUE)
se
```

But here we use the tibble view.

```{r}
options("restore_SummarizedExperiment_show" = FALSE)
se
```

We can now use tidyverse commands to interact with the
`SummarizedExperiment` object.

Podemos utilizar `filter` para filtrar as linhas utilizando uma condição, por exemplo, para visualizar
todas as linhas de uma amostra.

```{r}
se |> filter(.sample == "GSM2545336")
```

We can use `select` to specify columns we want to view.

```{r}
se |> select(.sample)
```

We can use `mutate` to add metadata info.

```{r}
se |> mutate(center = "Heidelberg University")
```

We can also combine commands with the tidyverse pipe `|>`. Por exemplo, poderíamos combinar `group_by` e `summarise` para obter o total de contagens para cada amostra.

```{r}
se |>
    group_by(.sample) |>
    summarise(total_counts=sum(counts))
```

We can treat the tidy SummarizedExperiment object as a normal tibble
for plotting.

Aqui traçamos a distribuição das contagens por amostra.

```{r tidySE-plot}
se |>
    ggplot(aes(counts + 1, group=.sample, color=infection)) +
    geom_density() +
    scale_x_log10() +
    theme_bw()
```

For more information on tidySummarizedExperiment, see the package
website
[here](https://stemangiola.github.io/tidySummarizedExperiment/).

**Take-home message**

- `SummarizedExperiment` represents an efficient way to store and
  handle omics data.

- They are used in many Bioconductor packages.

Se seguir a próxima formação centrada na análise de sequências de RNA, aprenderá a utilizar o pacote Bioconductor `DESeq2` para efetuar algumas análises de expressão diferencial.  Toda a análise do pacote `DESeq2`
é tratada num `SummarizedExperiment`.

:::::::::::::::::::::::::::::::::::::::: keypoints

- Bioconductor is a project provide support and packages for the
  comprehension of high high-throughput biology data.
- A `SummarizedExperiment` is a type of object useful to store and
  manage high-throughput omics data.

::::::::::::::::::::::::::::::::::::::::::::::::::
